---
title: "Code-based open-source software for Teaching Interactive Data Visualisation"
subtitle: "Mid Semester 1 Report"
author: "Shan-I Lee"
runtime: shiny
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(knitr)
library(plotly)
library(shiny)
library(crosstalk)
library(lattice)
library(dplyr) 
load("AQ.RData")
```

## Introduction
This report will outline a possible structure for introducing graduate students to interactive data visualisation techniques. A possible sequence for instruction will be discussed, followed by examples of applying interactive techniques with the open-source tools researched so far. The challenges posed by large datasets will also be examined. Areas in need of further research will be identified throughout the report and then summarised.


### A sequence for instruction
The data analysis process described by Cook and Swayne (2007) provides a structure and context for meaningful interactive visualisation. They describe how visualisation fits into the process of identifying the problem statement, preparing the data, engaging in exploratory data analysis (EDA) and quantitative analysis, and lastly in the presentation of findings. Graduate students will have encountered the use of static visuals in exploratory data analysis, quantitative analysis (typically for the purpose of checking model assumptions) and the use of graphics in presenting findings. However the role of visualisation in cleaning data may be less familiar. In particular the use of interactive visuals to identify and explore missing values, as described by Cook and Swayne (2007, Chapter 3), could provide a meaningful context for students to practice their interactive techniques and tools.
It is proposed that students initially practice interactive tools and techniques with an EDA frame of mind, but with more expertise explore specific applications, such as dealing with missing values or exploring cluster analysis visually. From the onset students need to be aware of how the interactive toolbox they are building can be of use throughout the data analysis process. A worked example demonstrating the use of interactive tools at different stages of the process, may assist in establishing this holistic view of interactive data visualisation. The airlines data from the 2009 ASA Data Expo could be used as a holistic and motivational exemplar before introducing interactive tools and techniques. A subset of the airlines data will be the main dataset used in this report, but the inclusion of other datasets for teaching will be addressed. Initial work on an introductory exemplar will be detailed, followed by a summary of the coverage of tools researched so far. A possible sequence for introducing the tools will then be outlined.


#### Choice of datasets
The examples in this report will be mainly drawn from the subset of flights by the A
Three data sets will be used for demonstration and practice, mtcars from the `R` datasets package, a data set of the fatty acid compositions of Italian olive oils and a subset of the airlines data from the 2009 ASA Data Expo. The data sets have been chosen for their size and combination of numeric and categorical variables. The familiarity of the mtcars' context and its small size, allows the learner to focus on experimenting with the interactive tool and technique at hand. Meanwhile the olive oils data set may provide more meaningful interaction as the relationships between variables are not as obvious. Furthermore insight into the Olive Oils data can be gained from interacting with bivariate scatterplots to complex multi-dimensional tours, as described in detail by Cook and Swayne. Lastly the subset of the airlines data comprising of more than 50 000 observations, is relatively small in the context of 'big data' but it will be large enough to demonstrate some of the challenges posed with 'scaling up' interactive tools.
Inclusion of a data set with only categorical vars? Interactive mosaic plot with changing nesting.
Airlines data is not cleaned yet.


#### An introductory exemplar
The purpose of the introductory exemplar is to provide students with a holistic view of interactive data visualisation. When learning the specific tools and techniques students can easily lose sight of the wider purpose of the interaction, other than 'making it work'. The context of the airlines data is easy to relate to and hence naturally motivates meaningful questions to investigate with the data. Furthermore the airlines data has enough depth for use as an exemplar, while leaving room for students to use again for further exploration. The suggested exemplar, below, uses the `AQ` subset of the airlines data. 

```{r, echo=FALSE}
radioButtons(inputId = "month", label = "Choose a Month (of 2001)",
               choices = c("January" = 1,
                           "February" = 2,
                           "March" = 3,
                           "April" = 4,
                           "May" = 5,
                           "June" = 6,
                           "July" = 7,
                           "August" = 8,
                           "September" = 9,
                           "October" = 10), inline = 1)

renderPlot({
      parallelplot(AQ[AQ$Month == input$month, 
                    c("Distance", "AirTime", "DepTime", "DayofMonth")],
      groups = AQ$DayOfWeek,
      auto.key = list(space = "right"),
      main = "AQ Carrier flights in 2001")
})
```

Below is an outline of how the exemplar fits into the data analysis process. 

The problem statement: How did the 9/11 terrorist attacks affect the AQ airline carrier?  
Investigating the broader problem statement of "How did the 9/11 terrorist attacks affect airlines?" is currently hindered by challenges with visually representing large datasets. This will be discussed as an area in need of further research.

Preparing data: We can see beyond January that there may have been inconsistencies in recording the day of the week with its respective day of the month. (We can also see a gap in the data in the days following September 11).

Exploratory data analysis: There appears to be two main structures for relationships between the departure time, air time and distance travelled. January, July and August appear to have a similar structure with shorter flights departing throughout the day. The remaining months have flights departing later in the day, apart from a few outliers. January, July and August correspond with the school holiday seasons in the US and hence suggests in the quantitative analysis, September should only be compared with the non-school-holiday months that are expected to be similar in structure. 

Quantitative analysis: One approach could be to build a model for a measure of airline efficiency such as the length of delays. `Month` could be included as a regressor and hence the adjusted effect of the month being September compared to other months of interest. This type of analysis would require diagnostics checks on multicollinearity. The linking of a correlation heatmap with scatterplots showcased by Sievert (2017) demonstrates how interaction can be used to overcome the difficulties of utilising a scatterplot matrix when there are a large number of variables (see <https://plot.ly/r/shinyapp-linked-click/>).   Hexbin plots have been used in place of scatterplots to reduce the delay in plotting time caused by the large number of observations in the `AQ` dataset.

```{r, echo=FALSE}
# Subset numeric vars
AQnumeric <- AQ[, c("DepTime", "CRSDepTime", "ArrTime", 
                    "CRSArrTime", "ActualElapsedTime", "CRSElapsedTime", 
                    "AirTime", "ArrDelay", "DepDelay", "Distance", 
                    "TaxiIn", "TaxiOut")]
# Correlation matrix
correlation <- round(cor(AQnumeric), 3)
nms <- names(AQnumeric)

renderPlotly({
  # Interactive correlation matrix as a heatmap w tooltips
  plot_ly(x = nms, y = nms, z = correlation,
          key = correlation, type ="heatmap", source = "heatplot") %>%
    layout(xaxis = list(title = ""),
           yaxis = list(title = ""))
})

renderPlotly({
  s <- event_data("plotly_click", source = "heatplot")
  # event_data() listens for the "selected" event in the "source"
  print(s) # Output from brush interaction
  if (length(s)){
    # Subset AQnumeric based on interaction
    d <- AQnumeric[, c(s$x, s$y)]
    p <- ggplot(d, aes(d[,1],d[,2] )) +
      geom_hex()
    ggplotly(p) %>%
      layout(xaxis = list(title = s[["x"]]),
             yaxis = list(title = s[["y"]]),
             showlegend = FALSE)
  } else {
    plotly_empty()
  }
})
```

The exemplar provides a brief glimpse on how interactive visualisation fits into the data analysis process. It should be emphasised that visuals from multiple views are required in analysis and the exemplar is not a complete exploration.  Instead it is intended for students to return to this data set to explore more complex interactive visualisations, such as multivariate cluster analysis. This possibility needs to be investigated further but it could follow a similar sequence as outlined by Cook and Swayne (2007) with the `Italian Olive Oils` data set. Furthermore students familiar with database querying may return to the problem statement with their interactive toolbox to analyse the data across the years. 

The visualisations shown in the exemplar demonstrate in a meaningful context the various interactive techniques that students will encounter. The exemplar might also be students' first encounter with a parallel coordinate plot. Although the focus is on interactive techniques, an introduction to useful, but less well known, multivariate visuals like parallel coordinate plots and mosaic plots may be required. An interactive visualisation using mosaic plots, with the categorical variables of the airlines data, could be developed (see Hofmann, 2006, Chapter 5). The introduction of mosaic plots could be discussed as one approach towards addressing the challenges of visualising datasets with a large number of categorical variables. 

The exemplar needs to be further developed and possibly extended, but the intention is for it to set the stage for applying interactive techniques in a meaningful context.


#### Introducing interactive tools 
The tables below summarise the interactive techniques described by Cook and Swayne (2007) covered by the `R` packages researched so far. Linked brushing is summarised separately since there is a variety of ways to brush and link plots. Further research is needed into other tools like `GGobi` and `ggvis`.  


###### Table 1: Interactive techniques for `Shiny`, `Plotly` and `Crosstalk`
```{r table1, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
tabl <- "
| R Package | Talks back to R | Identification | Scaling | Subset selection | Drag points | Brushing |
|-----------|:---------------:|:--------------:|:-------:|:----------------:|:-----------:|:--------:|
| Shiny | Y | | | Y<br>(input objects) | | | 
| Plotly | | Y<br>(tooltips) | Y | Y<br>(legend icons) | | Y |
| Crosstalk &<br>htmlwidgets | Y | Y | Y | Y<br>(filtering) | Y<br>(D3) | Y |
"
cat(tabl)
```


###### Table 2: Linked brushing techniques
```{r table2, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
tabl2 <- "
| R Packages | Aggregate<br>brushing | Persistent<br>brushing | Brush lines | Brush points<br>(link 1-to-1) | Brush to<br>link n-to-m |
|------------|:---------------------:|:----------------------:|:-----------:|:-----------------------------:|:-----------------------:|
| Plotly+Crosstalk |  | Y |  | Y |  | 
| Plotly+Shiny | Y | Y |  | Y |  |
"
cat(tabl2)
```

Of all the `R` packages listed above, graduate students are most likely to have had experience with using `Shiny`. Students who have not encountered `Shiny` before can be directed to its self-explanatory tutorials (<https://shiny.rstudio.com/tutorial/>).  In particular *Part 1 - How to build a `Shiny` app* familiarises students with the code structure of `Shiny` apps. The video tutorial for *Part 2 - How to customize reactions*, could be used as part of a 'flipped classroom' approach towards learning more advanced aspects of designing `Shiny` apps.  A hands-on face-to-face tutorial following the viewing of the online material for Part 2, may enable students to pick up the material faster than they would independently. The in-lecture tutorial would also act as a checkpoint for students to be on a par with the level of expertise required to proceed with utilising `Shiny` in combination with other packages. More guidance will be required for less thoroughly resourced packages. The showcase examples aim to 'show off' the capabilities of a package and hence can be more complex than needed for a novice. 

The plotly package provides a starting point that leads to a variety of interactive techniques. The default settings for a plotly object allows for tooltip identification and scaling. Students have the option of converting plots in `ggplot2` to plotly objects or directly creating plots using `plot_ly()`. The `plot_ly` function takes a similar layered Grammar of Graphics (Wilkinson 2005) approach as `ggplot2`. Although the graphics of `ggplot2` are more comprehensive, creating plots via `plot_ly()` should also be utilised since it interacts more directly with the `plotly.js` library (Sievert 2017).

The `tooltip` argument for plotly objects allows for customisation of the identification information to display when hovering on a graphical feature. The possibility of including identification information other than those visible in the aesthetics of the plot, can be explored further by examining the `layerData` and `originalData` arguments of plotly objects.

Continuing onto combining `Plotly` with `Crosstalk` introduces the key interactive technique of brushing to link multiple plots. This can also be achieved by using `Plotly` and `Shiny` together, but the `SharedData` environment of the `crosstalk` package directly allows brushing to be initiated in any of the linked plots by default. However the `crosstalk` package is not appropriate for use with large data sets and currently supports only linked brushing and filtering views, for certain HTML widgets where observations are individually displayed (see <https://rstudio.github.io/crosstalk/>).  Further research into how the `htmlwidgets` package can be used with `Crosstalk`, with or without the `reactive` function in `Shiny`, is still required. The use of `D3` graphics will enable the interactive technique of dragging points, but how this is utilised in the data analysis process needs further research.  Linking with `Leaflet` maps would also be of interest, since like many other real-life datasets, the `airlines` data lends itself to a geospatial representation. Examining the use of `Crosstalk` with `htmlwidgets` would provide more varied and in-depth practice of utilising the `SharedData` environment for students.

The following example of using `Plotly` and `Crosstalk` together demonstrates how persistent brushing can be easily incorporated by an argument in the `plotly` function `highlight`.

```{r, echo=FALSE}
# Subset AQ data since Crosstalk is not appropriate for large datasets
Jan <- AQ[AQ$Month=="1",]
Jan$rownames <- rownames(Jan)

sd <- SharedData$new(Jan, key=~rownames) 

p1 <- plot_ly(Jan, x=~AirTime, y=~NetDelay) %>%
  add_markers(alpha=0.1, color=I("black")) %>%
  # Layers on selected points 
  add_markers(data=sd)

p2 <- plot_ly(Jan, x=~DepTime, y=~ArrTime) %>%
  add_markers(alpha=0.1, color=I("black")) %>%
  add_markers(data=sd)

# Arranges plots side-by-side
subplot(p1, p2, titleX = TRUE, titleY = TRUE) %>%
  layout(dragmode="select") %>%
  hide_legend() %>%
  highlight(off="plotly_deselect", color= "blue", persistent=TRUE) 
# Note: persistent arguement in highlight()
```


Utilising the `plotly_click` mode in the `event_data` function for plotly objects in `shiny` apps allows for linked brushing on area plots of categorical variables. 

```{r, echo=FALSE}
AQ$Month <- as.factor(AQ$Month)

# Barchart that will be brushed 
renderPlotly({
  AQ %>% count(Month) %>%
  plot_ly(x = ~Month, y = ~n, colour = I("blue"), source = "barchart")
})
# Link to scatterplot
renderPlotly({
  s <- event_data("plotly_click", source = "barchart")
  print(s) 
  if (length(s)) {
    # Subset AQ using selection
    d <- AQ[AQ$Month==s$x, ]
    plot_ly(d, x = ~AirTime, y = ~NetDelay) %>%
      add_markers() 
  } else {
    plotly_empty()
  }
})
```

Students should be able to identify a similar code structure between this example and the linked brushing between the correlation heatmap and pairwise scatterplots encountered in the exemplar.  It may be possible to brush aggregate plots of real-valued variables with further research into how `plotly` captures traces for layers of `ggplot2` plots, as described by Sievert in [Extending ggplotly()](https://cpsievert.github.io/plotly_book/extending-ggplotly.html). 


### Large datasets
As previously mentioned the `AQ` subset of the airlines data is relatively small in the context of 'big data', but with over 50 000 observations and 20 variables, it provides students with some exposure to the challenges of interactive visualisations with large datasets. A different database query of the airline data would easily extend the size beyond the `AQ` subset. 

Datasets can be large in terms of the number of observations, or the number of variables involved, or both. The challenges of displaying visuals with a large number of observations will be discussed first, followed by possible student experiences of interactive visualisations involving a large number of variables. 


#### Datasets with a large number of observations
The challenges posed by datasets with a large number of observations become quickly apparent with the airlines data. The parallel coordinate plot in the exemplar demonstrates the problem of overplotting. The delay in interactivity between the heatmap and scatterplot highlights the sluggishness that results from plotting a large number of points on the screen.  The parallel coordinate plot uses interaction to create conditional plots and hence avoids the issue of plotting time. Unwin (2006) argues plotting a sample of a large dataset is insufficient for identifying errors, outliers or local features. Similarly Hofmann (2006) emphasises, as with any visualisation, we need to represent both the global and local features of large data sets.  Visually representing a large number of observations individually requires more research into `OpenGL` environments.  

A discussion with students regarding graphical formats would also entail from working with large data sets. The scalable property of object orientated formats is desirable, but impractical for storage when a visual contains a large number of points. Hence the use of raster images is the more efficient choice, since mapping to pixels is unaffected by the number of points involved (Unwin 2006). The importance of choosing an appropriate graphical format can be easily demonstrated by comparing the run time required in `R` to export any scatterplot of the `AQ` airline data, as a PDF and PNG file (the image consists of over 50 000 points).


#### Datasets with a large number of variables
The number of real-valued variables in the airlines data opens the possibility of introducing students to the use of multi-dimensional tours, as demonstrated by Cook and Swayne (2006) with the `Italian Olive Oils` dataset using GGobi. Further research would be needed to confirm whether the biplot function in `R`, combined with either a `Shiny` slider or a `htmlwidgets` frame, would be able to facilitate tours. Otherwise students could use GGobi, not only to experience an interactive visualisation of joint distributions, but also for exposure to open-source software that supports both interactive graphics and data analysis. The `R` packages discussed so far provide "pipe" components, but do not have the capabilities of a "pipeline" to facilitate data-to-plot and more importantly the inverse, plot-to-data updates, after visual interactions (Wickham et al. 2007). 


### Areas in need of further research 



## Conclusion
An overview of how interactive visualisation fits into the underlying data analysis process could be demonstrated with 
It also provides a way to understand the options that are created by utilising combinations of the packages together. More guidance may be required for becoming familiar with packages that allow for on-plot interactions, while self-directed learning should be encouraged for off-plot interactive tools. 
Balance between Focus on the techniques and purposes of interaction. The tools used to achieve them are likely to change, but still necessary to pick up in order to engage in data visualisation.
The challenge of focusing on the uses of interactive visualisation for data analysis rather than producing some thing to showcase a particular software tool.