---
title: "2D Projection Pursuit functions"
author: "Shan-I Lee"
date: "7/10/2017"
output: html_document
---

```{r setup, echo=TRUE}
knitr::opts_chunk$set(include = FALSE)
library(ggplot2)
library(plotly)
library(tourr)
library(abind)
library(shiny) # For (shiny+plotly) function
library(GGally) # For PCP using ggparcord()
library(colorspace) # For colour as an argument
library(crosstalk) # For (crosstalk+plotly) function
library(htmltools)
library(MASS) # For crabs dataset
data("crabs")
```

### Function with Shiny app
```{r shiny}
# dataset should have rownames as appropriate ID label for each obs
# pp_index takes the same values as the index arguement in guided_tour
# cmass, holes, (To do: lda_pp)
pp2D_shiny <- function (dataset, index="cmass", group=NULL,...) {
  # Subset real-valued vars (type="double") for touring
  num_cols <- sapply(dataset, typeof)=="double"
  Xdataset <- dataset[, num_cols]
  # Save tour
  if (index=="cmass") {
    t <- save_history(Xdataset, guided_tour(cmass, d=2, max.tries = 50), sphere = TRUE, max=50,...) 
  } else if (index=="holes") {
    t <- save_history(Xdataset, guided_tour(holes, d=2, max.tries = 50), sphere = TRUE, max=50,...)
  } else {
    stop("Invalid 'index' argument. Choose either: cmass or holes")
  }
  # Start view as orthogonal projection of first two measures
  # Retain orthogonal projection as initial basis
  t0 <- t[,,1]
  t0 <- matrix(c(1, rep(0, dim(Xdataset)[2]), 1, rep(0, (dim(Xdataset)[2]-2))), ncol = 2)
  t1 <- abind(t0, t, along = 3)
  class(t1) <- "history_array"
  tinterp <- interpolate(t1) 
  #tinterp <- interpolate(t) # Uses bases to create smooth tour
  X_sphere <- attr(t, "data") #(n by p) 
  # Central mass index and projection pursuit tour function
  cmass_tour <- function(basis) {
    # Projected data matrix 
    XA <- X_sphere%*%matrix(basis, ncol=2) # (n by d)
    cmass_index <- (sum(exp(-0.5*diag(XA%*%t(XA))))/dim(XA)[1]-exp(-dim(XA)[2]/2))/(1-exp(-dim(XA)[2]/2))
    list(rescale(XA), cmass_index)
  }
  # Holes index
  holes_tour <- function(basis) {
    XA <- X_sphere%*%matrix(basis, ncol=2) # (n by d)
    holes_index <- (1-sum(exp(-0.5*diag(XA%*%t(XA))))/dim(XA)[1])/(1-exp(-dim(XA)[2]/2))
    list(rescale(XA), holes_index)
  }
  # Apply index function to each projection basis
  pp_index <- data.frame(iteration=1:length(tinterp))
  # 3rd component of tinterp is (1 by 2p) containing x-coeffs for each measurement var followed by y-coeffs
  # matrix(3rd_comp, ncol=2) converts it into a projection matrix, A.
  if (index=="cmass") {
    t_tour <- apply(tinterp, 3, FUN = cmass_tour)
  } else if (index=="holes") {
    t_tour <- apply(tinterp, 3, FUN = holes_tour)
  }
  
  for(i in 1:length(t_tour)) {
    pp_index$index[i] <- unlist(t_tour[[i]][2])
  }
  # Colour by Group 
  if (length(group)) {
    fac <- unlist(subset(dataset, select = group)) %>% unlist() %>% as.factor()
    pal <- rainbow_hcl(length(levels(fac))) # Compute a rainbow of colours (qualitative palette)
    colour <- as.factor(pal[as.numeric(fac)])
    col <- data.frame(col=colour, group=fac)
  } else {
    col <- data.frame(col="black", group="NA")
  }
  
  # Shiny app ---------------------------------------------------------------
  
  ui <- fluidPage(navbarPage("2D Guided Tour",
                             tabPanel("Animation",
                                      mainPanel(
                                        sliderInput("iteration", "Iteration", min = 1, value = 1, step = 1,
                                                    max = length(t_tour), animate = animationOptions(400), width = 500),
                                        plotlyOutput("tour")
                                      )       
                             ),
                             tabPanel("Proj Pursuit Index",
                                      splitLayout(
                                        plotlyOutput("index"),
                                        plotlyOutput("tour2")
                                      )
                             )
  ))
  
  server <- function(input, output, session) {
    output$tour <- renderPlotly({
      proj <- as.data.frame(t_tour[[input$iteration]][[1]])
      proj$ID <- rownames(dataset)
      p <- ggplot(proj, aes(x=V1, y=V2, label=ID, col=col$col, label1=col$group)) +
        geom_point() +
        scale_x_continuous(limits = c(-0.1, 1.1)) +
        scale_y_continuous(limits = c(-0.1, 1.1)) +
        theme_void() +
        theme(legend.position = "none")
      tour_p <- ggplotly(p, tooltip = c("label", "label1")) 
      basis <- as.data.frame(matrix(tinterp[[input$iteration]], ncol = 2))
      basis$measure <- colnames(X_sphere)
      basis$magnitude <- (basis$V1)^2+(basis$V2)^2
      axis_p <- ggplot(basis, aes(x=V1, y=V2)) +
        geom_segment(aes(xend=0, yend=0, col=magnitude)) +
        geom_text(aes(label=measure)) +
        scale_x_continuous(limits = c(-1.1, 1.1)) +
        scale_y_continuous(limits = c(-1.1, 1.1)) +
        theme_void() +
        theme(legend.position = "none") +
        scale_color_gradient(low = "#56B1F7", high = "#132B43")
      axis_p <- ggplotly(tooltip = "none")
      subplot(tour_p, axis_p)
    })
    
    output$index <- renderPlotly({
      ggplot(pp_index, aes(x=iteration, y=index)) +
        geom_point() +
        geom_line() +
        ggtitle("Projection pursuit index")
      ggplotly(source = "index_plot")
    })
    
    output$tour2 <- renderPlotly({
      s <- event_data("plotly_click", source = "index_plot")
      #print(s)
      if (length(s)) {
        iter <- s$x
      } else {
        iter <- 1
      }
      proj <- as.data.frame(t_tour[[iter]][[1]])
      proj$ID <- rownames(dataset)
      p <- ggplot(proj, aes(x=V1, y=V2, label=ID, col=col$col, label1=col$group)) +
        geom_point() +
        scale_x_continuous(limits = c(-0.1, 1.1)) +
        scale_y_continuous(limits = c(-0.1, 1.1)) +
        theme_void() +
        theme(legend.position = "none")
      tour_p <- ggplotly(p, tooltip=c("label", "label1")) 
      basis <- as.data.frame(matrix(tinterp[[iter]], ncol = 2))
      basis$measure <- colnames(X_sphere)
      basis$magnitude <- sqrt((basis$V1)^2+(basis$V2)^2)
      axis_p <- ggplot(basis, aes(x=V1, y=V2)) +
        geom_segment(aes(xend=0, yend=0, col=magnitude)) +
        geom_text(aes(label=measure)) +
        scale_x_continuous(limits = c(-1.1, 1.1)) +
        scale_y_continuous(limits = c(-1.1, 1.1)) +
        theme_void() +
        theme(legend.position = "none") +
        scale_color_gradient(low = "#56B1F7", high = "#132B43")
      axis_p <- ggplotly(tooltip = "none")
      subplot(tour_p, axis_p)
    })
  }
  shinyApp(ui, server)
}
```

#### Testing: 
```{r test_shiny}
pp2D_shiny(crabs)
pp2D_shiny(crabs, group = "sex")
pp2D_shiny(crabs, index = "holes", group = "sex")
```

#### Findings for NCEA data
```{r ncea, include=FALSE}
# For wrangling data
library(dplyr)
library(tidyr)
nzqa2016 <- read.csv("http://www.nzqa.govt.nz/assets/Studying-in-NZ/Secondary-school-and-NCEA/stats-reports/2016/Qualification-Statistics-School-2016-29032017.csv")
# Drop vars that will not be used (eg. cumulative achievement)
nzqa <- nzqa2016[,-c(1,8,10)]
names(nzqa) <- c("Decile", "Region", "School", "Year", "Qualification", "Achieve_participate", "Achieve_roll", "Small_sch")
levels(nzqa$Qualification) <- c("L1", "L2", "L3", "UE")
# Subset to use only Year 11 with Level 1, etc
nzqa <- nzqa %>% filter(((Qualification=="L1")&(Year==11))|((Year==12) & (Qualification=="L2"))|
                          ((Year==13) & (Qualification=="L3"))|((Year==13) & (Qualification=="UE")))
# Reshape so that one row = one school
# Current.Achievement.Rate.Participation kept for analysis only
achieved <- nzqa %>% 
  spread(Qualification, Achieve_participate, fill=0) %>%
  group_by(School) %>%
  summarise_at(c("L1", "L2", "L3", "UE"), sum) %>%
  inner_join(nzqa[, c(1, 2, 3, 8)]) %>% #Add Decile and Region and Small_sch variables
  distinct() %>% #One row per school
  filter(!((L1==0)&(L2==0)&(L3==0))) #Remove schools with 0% achievement rate for all levels
# Function to replace 0% with NA
zeros <- function(col) {
  replace(col, col==0, NA)
}
achieved$L1 <- zeros(achieved$L1)
achieved$L2 <- zeros(achieved$L2)
achieved$L3 <- zeros(achieved$L3)
achieved$UE <- zeros(achieved$UE)
# Remove schools with 'small cohort' warning (obscures pattern in non-small cohorts).
achieved <- achieved[achieved$Small_sch=="",] #437 school left
achieved <- achieved[, -8]
# Remove obs with any NA values (Otherwise errors)
ach_narm <- achieved[complete.cases(achieved),] #407 schools left
rownames(ach_narm) <- ach_narm$School
ach_narm$Decile <- as.factor(ach_narm$Decile)
# Run function
pp2D_shiny(ach_narm, index = "cmass", group = "Decile")
pp2D_shiny(ach_narm, index = "cmass", group = "Region")
pp2D_shiny(ach_narm, index = "holes", group = "Decile")
pp2D_shiny(ach_narm, index = "holes", group = "Region")
```

##### Central mass index (Redo! Refer to PCs not X-vars directly)
A guided tour optimising the central mass index shows the concentration of decile 10 schools at the centre of the plot. Schools from the remaining deciles are scattered around the central mass along with a handful of decile 10 schools. 
The central mass index tended to be optimised when `UE` and one other variable was weighted heavier than the remaining variables. (The "other variable" varied between tours).
    + Double check interpretation with literature/theory 
    + Is this consistent with the findings from the supervised cluster analysis? (clust6)
        - Decile 10 schools part of the bigger of the two clusters.
        - Schools in the bigger cluster were more consistent with each other than the smaller cluster.
    + Replicate the `pp2D_fns` so that the user can compare the outcomes of repeated tours and compare for consistency.

The higher proportion of Auckland region schools compared to the other regions might be contributing to the visual effect of Auckland schools being at the centre when the central mass index is optimised.
    + Use opaque colours to address overplotting.
    + Increase tour plot height and width.
    
##### Holes index (Not so useful for NCEA data)
Optimising the holes index does not appear to provide any useful views. It takes at most a few distinct bases for the index to be optimised and the projections suggest there is no clear separation between schools of different decile or region. 

### Function with Crosstalk
```{r xtalk, echo=FALSE}
pp2D_xtalk <- function(dataset, index="cmass", ...) {
  # Subset real-valued vars (type="double") for touring
  num_cols <- sapply(dataset, typeof)=="double"
  Xdataset <- dataset[, num_cols]
  p <- length(Xdataset) # Number of real-valued Xs used in tour
  # Save tour
  if (index=="cmass") {
    t <- save_history(Xdataset, guided_tour(cmass, d=2, max.tries = 50), sphere = TRUE, max=50,...) 
  } else if (index=="holes") {
    t <- save_history(Xdataset, guided_tour(holes, d=2, max.tries = 50), sphere = TRUE, max=50,...)
  } else {
    stop("Invalid 'index' argument. Choose either: cmass or holes")
  }
  # Start view as orthogonal projection of first two measures
  # Retain orthogonal projection as initial basis
  t0 <- t[,,1]
  t0 <- matrix(c(1, rep(0, p), 1, rep(0, (p-2))), ncol = 2)
  t1 <- abind(t0, t, along = 3)
  class(t1) <- "history_array"
  tinterp <- interpolate(t1) 
  X_sphere <- attr(t, "data") #(n by p) 
  # Central mass index and projection pursuit tour function
  cmass_tour <- function(basis) {
    # Projected data matrix 
    XA <- X_sphere%*%matrix(basis, ncol=2) # (n by d)
    cmass_index <- (sum(exp(-0.5*diag(XA%*%t(XA))))/dim(XA)[1]-exp(-dim(XA)[2]/2))/(1-exp(-dim(XA)[2]/2))
    list(rescale(XA), cmass_index)
  }
  # Apply index function to each projection basis
  t_tour <- apply(tinterp, 3, FUN = cmass_tour)
  pp_index <- data.frame(iteration=1:length(tinterp))
  x <- c()
  y <- c()
  PC_x <- c()
  PC_y <- c()
  for(i in 1:length(t_tour)) {
    pp_index$index[i] <- unlist(t_tour[[i]][2])
    x <- c(x, unlist(t_tour[[i]][[1]][,1]))
    y <- c(y, unlist(t_tour[[i]][[1]][,2]))
    # Take first 'p' values to be x-coords for PCs
    PC_x <- c(PC_x, unlist(tinterp[[i]][1:p])) 
    # Take remaining 'p' values to be y-coords for PCs
    PC_y <- c(PC_y, unlist(tinterp[[i]][(p+1):(p+p)])) 
  }
  proj <- data.frame(Name=rep(rownames(dataset), length(tinterp)), 
                   step=rep(1:length(tinterp),each=dim(Xdataset)[1]), x=x, y=y)
  basis <- data.frame(measure=rep(colnames(X_sphere), length(tinterp)),
                   step=rep(1:length(tinterp),each=p), 
                   magnitude=PC_x^2+PC_y^2, x=PC_x, y=PC_y)
  
  # Axes for tour plot ("tour") and measurement var coeffs plot ("axes")
tx <- list(
  title = "", range = c(-0.1, 1.1), 
  zeroline = F, showticklabels = F
)

ax <- list(
  title = "", range = c(-1.1, 1.1), 
  zeroline = F, showticklabels = F
)
  # tour plot
  tour <- proj %>%
  SharedData$new(~Name, group = "2Dtour") %>%
  plot_ly(x = ~x, y = ~y, frame = ~step, color = I("black"), 
          height = 450, width = 800) %>%
  add_markers(text = ~Name, hoverinfo = "text") %>%
  layout(xaxis = tx, yaxis = tx)
  
  axes <- basis %>%
  plot_ly(x = ~x, y = ~y, frame = ~step, hoverinfo = "none") %>%
    add_segments(xend = 0, yend = 0, color = I("gray85"), size = I(1)) %>%
    # Color segments don't work
  #add_segments(xend = 0, yend = 0, color = ~magnitude, size = I(1), showlegend=FALSE) %>%
  add_text(text = ~measure, color=I("black")) %>%
  layout(xaxis = ax, yaxis = ax)
  # Doesn't work if use ggplot() for axes plot
  
  # very important these animation options are specified _after_ subplot()
# since they call plotly_build(., registerFrames = T)
# (Took a few minutes to run)
  tour <- subplot(tour, axes, nrows = 1, margin = 0) %>%
  animation_opts(33) %>% #33 milliseconds between frames
  hide_legend() %>%
  layout(dragmode = "select") %>%
  highlight(on="plotly_select", off= "plotly_deselect", color = "blue", persistent = TRUE)
  
  html <- tags$div(
  style = "display: flex; flex-wrap: wrap",
  tags$div(tour, align = "center", style = "width: 50%; padding: 1em;")
  #tags$div(brush_group, align = "center", style = "width: 50%; padding: 1em;")
)

# opens in an interactive session
res <- html_print(html)
}
```

#### Testing: 
```{r test_xtalk}
pp2D_xtalk(crabs)
pp2D_xtalk(crabs, index = "holes")
```
